---
title: "13: High Frequency Data"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Determine stormflow and baseflow from high frequency flow data
2. Use a hysteresis plot to understand watershed dynamics
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

High frequency data is usually defined as frequencies significantly lower than daily (e.g. 5-minute, 15-minute, 1 hr etc). What types of hydrological and biological processes happen on this timescale that we might want to investigate?

> Short lived storms can have a big impact but might not be caught by daily data. Contaminant transport, and other things that have big daily changes, like temperature and dissolved oxygen.


## Session Set Up

```{r, message = FALSE}
getwd()

packages <- c(
  "dataRetrieval", "tidyverse", "cowplot", "EcoHydRology", "xts", "dygraphs"
  )
invisible(lapply(packages, library, character.only = TRUE)) 

theme_set(theme_classic(base_size = 12))
```

## Pull Data

```{r download data}

## use readNWISuv for instantaneous values (anything more frequent than daily)
## daily values would take the mean of the instantaneous values
## time is all in UTC (greenwich)
dischargedatTFC2018 <- readNWISuv(
  site = "02097280", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()


dischargedatEC2018 <- readNWISuv(
  site = "0208675010", #Ellerbe Creek at Club Boulevard at Durham, NC
  parameterCd = "00060", 
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()


plot_grid(
  ggplot(dischargedatTFC2018, aes(x = dateTime, y = Flow_Inst)) +
    geom_line() + 
    scale_y_log10() +
    ggtitle("Third Fork Creek"),
  ggplot(dischargedatEC2018, aes(x = dateTime, y = Flow_Inst)) + 
    geom_line() + 
    scale_y_log10() +
    ggtitle("Ellerbe Creek"),
  ncol = 1
)
          

```


The shape and size of river hydrographs following rain events can tell us how water moves through a watershed. Once rain falls, how does water get to a stream? What would the streamflow do?

> Overland flow - either through pipes or "normal" overland flow. Percolation to groundwater and eventually to the river. Direct precipitation in the stream.

> Baseflow generally follows the troughs of a hydrograph.


_"Hydrograph separation is one of the most desperate analysis techniques in use in hydrology" - Hewlett & Hibbert 1967_


```{r baseflowseparation}
TFCbaseflow <- BaseflowSeparation(
  dischargedatTFC2018$Flow_Inst, 
  filter_parameter = 0.925, # default parameter
  passes = 3 # default parameter
  )

# creates baseflow tme series and quick flow time series

## Now need to add back in time stamps with cbind()

TFC2018 <- cbind(dischargedatTFC2018, TFCbaseflow)

ggplot(TFC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
  # scale_y_log10() +
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(TFC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(TFC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(TFC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()
## quickflow is the majority of streamflow in the early part of the streamflow increase
## baseflow is the majority of the flow after the overall streamflow peaks


Export <- TFC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_), # difference in seconds, with NA appended at the end
         baseflowexport = bt * timestep, # cubic feet
         quickflowexport = qft * timestep) %>% # cubic feet
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)

percentbf <- Export$BaseflowExport_cf/Export$TotalExport_cf
percentqf <- Export$QuickflowExport_cf/Export$TotalExport_cf
```

What percentage of water is exported from this watershed as baseflow? What percentage as quickflow?

> 51.9% baseflow, 48.1% quickflow.

Repeat this analysis for Ellerbe Creek. How do the percentages compare?

```{r}
ECbaseflow <- BaseflowSeparation(
  dischargedatEC2018$Flow_Inst, 
  filter_parameter = 0.925, # default parameter
  passes = 3 # default parameter
  )

# creates baseflow tme series and quick flow time series

## Now need to add back in time stamps with cbind()

EC2018 <- cbind(dischargedatEC2018, ECbaseflow)

ggplot(EC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
  # scale_y_log10() +
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

dygraph(
  cbind(
    Flow = with(EC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(EC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(EC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()
## quickflow is the majority of streamflow in the early part of the streamflow increase
## baseflow is the majority of the flow after the overall streamflow peaks


ExportEC <- EC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_), # difference in seconds, with NA appended at the end
         baseflowexport = bt * timestep, # cubic feet
         quickflowexport = qft * timestep) %>% # cubic feet
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)

percentbfEC <- ExportEC$BaseflowExport_cf/Export$TotalExport_cf
percentqfEC <- ExportEC$QuickflowExport_cf/Export$TotalExport_cf
```


> 43.4% baseflow, 56.6% quickflow.

What does this mean physically about the watershed?

> The greater amount of quickflow in Ellerbee creek likely means there are more impervious surfaces than in the Third Fork Creek watershed. Both however are urban streams.

## Hydrologic Flashiness

A "flashy" stream increases in flow faster following a precipitation event. 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> Flashiness should increase with urbanization and impervious surface development.

What do you think "urban stream syndrome" refers to?

> Flashy streams that that becme incised over time. Water quality parameters will also accompany urban stream syndrome.

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as $$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$

It is the sum of the differences in streamflow divided by the sum of flows at each time step, all divided by catchment area. Technically the Richar-Baker index does not normalize by catchment area.

```{r Flashiness}
TFCsitedata <- readNWISsite(site = "02097280")
TFC.catchment.size <- TFCsitedata$drain_area_va #square miles

RBI_TFC2018 <- (
  sum(abs(diff(dischargedatTFC2018$Flow_Inst))) / 
    sum(dischargedatTFC2018$Flow_Inst[-1])
  ) / TFC.catchment.size

ECsitedata <- readNWISsite(site = "0208675010")
EC.catchment.size <- ECsitedata$drain_area_va #square miles

RBI_EC2018 <- (
  sum(abs(diff(dischargedatEC2018$Flow_Inst))) / 
    sum(dischargedatEC2018$Flow_Inst[-1])
  ) / EC.catchment.size

```

Which stream is flashier?

> Ellerbee creek is the flashier creek.

Repeat the process for a stream of our choice

```{r}
## First look for sites with the desired data
sites <- whatNWISsites(parameterCD = "00060", statecd = "NC")
View(sites)

## check if the site has continuous data
LTRdata <- whatNWISdata(site = "03517533")

## Choosing LITTLE TENNESSEE RIVER AT TAPOCO, NC, site_no = 03517533
dischargedatLTR2018 <- readNWISuv(
  site = "03517533", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()

```


## Chemical Flashiness

Chemical concentrations in rivers can be either highly variable with discharge, or "chemostatic", which means the concentration barely changes with discharge. 

```{r}
CCdat <- readNWISuv(
  site = "02249500", # CRANE CREEK AT MELBOURNE, FL
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2015-01-1",
  endDate = "2016-01-01"
  ) %>%
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(CCdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


Potdat <- readNWISuv(
  site = "01646500", # POTOMAC RIVER NEAR WASH, DC LITTLE FALLS PUMP STA
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2016-01-1",
  endDate = "2017-01-01"
  ) %>% 
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(Potdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Which of these rivers is more chemostatic?

>

## Hysteresis Loops

Hysteresis occurs when the concentration varies with discharge in a different relationship on the rising and falling limbs of a storm. This can be caused by _flushing_ behavior or _dilution_ behavior. 

```{r}
dygraph(
  cbind(
    Flow = xts(Potdat$Flow_Inst, order.by = Potdat$dateTime), 
    Nitrate = xts(Potdat$Nitrate_mgl, order.by = Potdat$dateTime)
    )
  ) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector()
```

Look at the storm that occured around February 1st. Do you expect this storm to be a flushing or a diluting storm?

>

```{r}
PotomacStorm <- Potdat %>%
  filter(dateTime > "2016-01-31" & dateTime < "2016-02-14") 

ggplot(PotomacStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() 


```

Did this storm exhibit clockwise or counterclockwise hysteresis?

>

Which direction do hysteresis loops go for diluting storms?

>

Which direction do hysteresis loops go for flushing storms?

>

## Closing Discussion

What hydrological processes did we investigate today? 

>

What can high-frequency data be used for that daily data may not be able to capture?

> 
